\texttt{escheR}: Optimize Spatial Visualization Following Gestalt Principles

### 2023-03-17 Manuscript/manuscript.tex
Replace escheR

%\setmainfont{Helvetica}
\title{\coloc: Optimize Spatial Visualization Following Gestalt Principles}
\author[1]{Boyi Guo\thanks{Corresponding author}}
\author[1]{Stephanie C. Hicks}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA}
\date{\today}

@@

%% ABSTRACT ============================================================================================
\section*{Abstract}
\subsection*{Summary}
Visualization is an indispensable component of data analysis and receives growing recognition as essential. Nevertheless, current methods and
tools are inadequate to visualize the novel spatially-resolved data, limiting the ability to derive new insights and discoveries. Specifically, there are no multi-dimensional \textit{in-situ} visualization that simultaneously displays multiple variables in a 2D spatial map. To address this problem, we introduce gestalt principles to the design of spatial maps, layering aesthetics to display multiple variables. The proposed visualization can be broadly applied to spatially resolved data and 2D embedding methods. We provide an open-source R package \coloc, which seamlessly integrates into the state-of-the-art spatial omics toolboxes.

\subsection*{Availability and implementation}
The R package \coloc is freely available at Github (\url{https://github.com/boyiguo1/escheR}).
% and Zenodo (\fixme{add link}).

@@

To implement the proposed visualization, we leverage the state-of-art data visualization framework \texttt{ggplot2}\cite{ggplot2} following the grammar of graphics\cite{wilkinson_2012}. Specifically, we use an interactive process to layer each variable in spatial maps using different aesthetics. For example, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map. We provide an open-source package called \coloc (named after the graphic artist M.C. Escher) in the R programming environment\cite{R}, providing a simplified interface to navigate the implementation of the multi-dimensional \textit{in situ} visualization. By adapting \texttt{ggplot2} standard, \coloc can be seamlessly integrated into many popular spatial resolved toolboxes, such as \texttt{SpatialLIBD}\cite{pardo_2022}, \texttt{Seurat}\cite{hao_2021}, \texttt{Giotto}\cite{dries_2021} to name a few, and allow further theme customization with ease. We provide two use cases to exemplify some utility of the proposed spatial visualization: 1) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex (DLPFC) using spatial transcriptomics data\cite{huukimyers_2023}; 2) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters \cite{freytag_2020}.

\section*{Use Case 1: Spatially Differential Colocalization in DLPFC}
In a recent study investigating the molecular organization of human dorsolateral prefrontal cortex (DLPFC) \cite{huukimyers_2023}, two schizophrenia risk genes, membrane-bound ligand ephrin A5 (\textit{EFNA5}) and ephrin type-A receptor 5 (\textit{EPHA5}), are identified to colocalize via the cell-cell communication analysis. In addition, data suggest Layer 6 is the most highly co-localized layer compared to other cortex layers. To visually examine the inference, we apply \coloc to create a multi-dimensional \textit{in-situ} spatial map (Figure \label{fig:visual}) that simultaneously exhibits the cortex layers (displayed with color-coded spot outlines) and the categorized colocalization status of \textit{FYN} and \textit{EFNA5} (displayed with color-coded spot fill). Compared to the traditional visualization where the cortex layers and the colocalization status are ploted in two side-by-side figures, our proposed visualization enables directly mapping colocalization status to the spatial domain, simplifying the perception of two sources of information and allowing cognitive comparison across cortex layers.

@@

\vspace{0.45in}
In summary, we propose an innovative multi-dimensional spatial visualization to simultaneously display multiple variables in a 2D coordinate system. Specifically, our design leverages the gestalt principle from visual perception to introduce other dimensions of a spatial map by iterative layering aesthetics. Developed upon \texttt{ggplot2}, we provide an open-source R package \coloc that is seamlessly compatible with popular spatial resolved data analysis toolboxes. The proposed visualization has broad applications in visualizing both spot-based and image-based spatial resolved data, spatial multi-omics as well as other spatial-based data. The proposed visualization innovation highlights the importance and potential of translating theories in vision science to data science, addressing long-standing challenges in bioinformatics data visualization.

% While \coloc is to design to display colocalization in their spatial context, the underlying principle can be broadly applied to address visualizaiton challenges in informaiton rich visualizations. The contrast color optical illusion provides allows to add another dimension of information in any spatial visualization, which can be further used to display multi-omics presence, greatly improve the interpretability.

@@

%% EPILOGUE ================================================================
\section*{Acknowledgements}
B.G. would like to acknowledge Dr. Leonardo Collado Torres, Louise Huuki-Myers and Leon Di Stefano for their helpful feedback.

>>>>

%\setmainfont{Helvetica}
\title{\texttt{escheR}: Optimize Spatial Visualization Following Gestalt Principles}
\author[1]{Boyi Guo\thanks{Corresponding author}}
\author[1]{Stephanie C. Hicks}
\affil[1]{Department of Biostatistics, Johns Hopkins Bloomberg School of Public Health, MD, USA}
\date{\today}

@@

%% ABSTRACT ============================================================================================
\section*{Abstract}
\subsection*{Summary}
Visualization is an indispensable component of data analysis and receives growing recognition as essential. Nevertheless, current methods and
tools are inadequate to visualize the novel spatially-resolved data, limiting the ability to derive new insights and discoveries. Specifically, there are no multi-dimensional \textit{in-situ} visualization that simultaneously displays multiple variables in a 2D spatial map. To address this problem, we introduce gestalt principles to the design of spatial maps, layering aesthetics to display multiple variables. The proposed visualization can be broadly applied to spatially resolved data and 2D embedding methods. We provide an open-source R package \texttt{escheR}, which seamlessly integrates into the state-of-the-art spatial omics toolboxes.

\subsection*{Availability and implementation}
The R package \texttt{escheR} is freely available at Github (\url{https://github.com/boyiguo1/escheR}).
% and Zenodo (\fixme{add link}).

@@

To implement the proposed visualization, we leverage the state-of-art data visualization framework \texttt{ggplot2}\cite{ggplot2} following the grammar of graphics\cite{wilkinson_2012}. Specifically, we use an interactive process to layer each variable in spatial maps using different aesthetics. For example, we use the combination of \texttt{color} and \texttt{fill="transparent"} to create the background layer and \texttt{fill} to create the figure layer. When necessary to display an additional layer for a third variable, \texttt{shape} can be used to add symbols such as cross (+) and asterisk (*) to highlight in the spatial map. We provide an open-source package called \texttt{escheR} (named after the graphic artist M.C. Escher) in the R programming environment\cite{R}, providing a simplified interface to navigate the implementation of the multi-dimensional \textit{in situ} visualization. By adapting \texttt{ggplot2} standard, \texttt{escheR} can be seamlessly integrated into many popular spatial resolved toolboxes, such as \texttt{SpatialLIBD}\cite{pardo_2022}, \texttt{Seurat}\cite{hao_2021}, \texttt{Giotto}\cite{dries_2021} to name a few, and allow further theme customization with ease. We provide two use cases to exemplify some utility of the proposed spatial visualization: 1) the spatially differential gene colocalization in the human dorsolateral prefrontal cortex (DLPFC) using spatial transcriptomics data\cite{huukimyers_2023}; 2) multi-dimensional UMAP highlighting differential gene expression in data-driven cell clusters \cite{freytag_2020}.

\section*{Use Case 1: Spatially Differential Colocalization in DLPFC}
In a recent study investigating the molecular organization of human dorsolateral prefrontal cortex (DLPFC) \cite{huukimyers_2023}, two schizophrenia risk genes, membrane-bound ligand ephrin A5 (\textit{EFNA5}) and ephrin type-A receptor 5 (\textit{EPHA5}), are identified to colocalize via the cell-cell communication analysis. In addition, data suggest Layer 6 is the most highly co-localized layer compared to other cortex layers. To visually examine the inference, we apply \texttt{escheR} to create a multi-dimensional \textit{in-situ} spatial map (Figure \label{fig:visual}) that simultaneously exhibits the cortex layers (displayed with color-coded spot outlines) and the categorized colocalization status of \textit{FYN} and \textit{EFNA5} (displayed with color-coded spot fill). Compared to the traditional visualization where the cortex layers and the colocalization status are ploted in two side-by-side figures, our proposed visualization enables directly mapping colocalization status to the spatial domain, simplifying the perception of two sources of information and allowing cognitive comparison across cortex layers.

@@

\vspace{0.45in}
In summary, we propose an innovative multi-dimensional spatial visualization to simultaneously display multiple variables in a 2D coordinate system. Specifically, our design leverages the gestalt principle from visual perception to introduce other dimensions of a spatial map by iterative layering aesthetics. Developed upon \texttt{ggplot2}, we provide an open-source R package \texttt{escheR} that is seamlessly compatible with popular spatial resolved data analysis toolboxes. The proposed visualization has broad applications in visualizing both spot-based and image-based spatial resolved data, spatial multi-omics as well as other spatial-based data. The proposed visualization innovation highlights the importance and potential of translating theories in vision science to data science, addressing long-standing challenges in bioinformatics data visualization.

% While \texttt{escheR} is to design to display colocalization in their spatial context, the underlying principle can be broadly applied to address visualizaiton challenges in informaiton rich visualizations. The contrast color optical illusion provides allows to add another dimension of information in any spatial visualization, which can be further used to display multi-omics presence, greatly improve the interpretability.

@@

%% EPILOGUE ================================================================
\section*{Acknowledgements}
B.G. would like to acknowledge Dr. Leonardo Collado Torres, Louise Huuki-Myers, Dr. Lukas W Weber, and Leon Di Stefano for their helpful feedback.
